White Wine Attributes Analysis by Amr Mohamed
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine_df <- read.csv('wineQualityWhites.csv')
```

# Dataset information
Throughout this report, the White Wine Quality dataset will be analyzed.<br>
The dataset contains 4,898 white wines with 11 variables on quantifying the 
chemical properties of each wine.<br>
At least 3 wine experts rated the quality of each wine,
providing a rating between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

Let's start by understanding the different attributes of the dataset.<br>
I will use basic histograms to discover the distributions of each attribute.

```{r echo=FALSE, message=FALSE, fig.align='center', Univariate_Plots}
ggplot(data = melt(subset(wine_df, select = -X)), mapping = aes(x = value)) + 
  geom_histogram(bins = 20, color ='black', fill=I('#328484')) +
  facet_wrap(~variable, scales = 'free_x')
```
```{r}
summary(wine_df)
```


The special distribution of the residual sugar encourages further investigation.<br>

```{r echo=FALSE, fig.align='center'}

high_sugar_wine_df <- subset(x = wine_df, subset = residual.sugar > quantile(x = wine_df$residual.sugar, probs = 0.9))
# ggplot(mapping = aes(x=quality), breaks=0:10) +
#   geom_histogram(data = high_sugar_wine_df, binwidth = 1,
#                  mapping = aes(y=..count../sum(..count..)),
#                  alpha=0.5, color ='black', fill=I('#FF8484'),
#                  show.legend=TRUE) +
#   geom_histogram(data = wine_df, binwidth = 1,
#                  mapping =aes(y=..count../sum(..count..)),
#                  alpha=0.5, color ='black', fill=I('#328484'),
#                  show.legend=TRUE) +
#   coord_cartesian(xlim = c(0, 10)) +
#   scale_x_continuous(breaks = seq(from=0, to = 10, by = 1)) +
#   ylab('Proportion') +
#   xlab('Quality of wine')

p1 <- ggplot(mapping = aes(x=quality, y=..count../sum(..count..)), breaks=0:10) +
  geom_histogram(data = high_sugar_wine_df, binwidth = 1,
                 alpha=0.5, color ='black', fill=I('#FF8484'),
                 show.legend=TRUE) +
  coord_cartesian(xlim = c(0, 10)) +
  scale_x_continuous(breaks = seq(from=0, to = 10, by = 1)) +
  ylab('Proportion') +
  xlab('Quality of wine') +
  ggtitle('General Distribution of wine\'s quality') + 
  theme(plot.title = element_text(size = 9, face = "bold", hjust = 0.5))

p2 <- ggplot(breaks=0:10) +
  geom_histogram(data = wine_df, mapping = aes(x=quality, y=..count../sum(..count..)), binwidth = 1,
                 alpha=0.5, color ='black', fill=I('#328484')) +
  scale_x_continuous(breaks = seq(from=0, to = 10, by = 1)) +
  coord_cartesian(xlim = c(0, 10)) +
  ylab('Proportion') +
  xlab('Quality of wine') +
  ggtitle('Distribution of High sugar residue wine\'s quality') + 
  theme(plot.title = element_text(size = 9, face = "bold", hjust = 0.5))
grid.arrange(p1, p2, nrow=1)
```

High sugar values doesn't seem to affect the quality. 

```{r echo=FALSE, results='hide'}
summary(wine_df$quality)
sd(wine_df$quality)
```

# Univariate Analysis

### Observations:
* The quality of wine has integral values.
* The alcohol attribute is long tailed and the distribution is nearly uniform.
* The residual sugar values has an exponentially decaying distribution.
* The distibution of wine's quality with higher residual sugar has similar distribution to the general distribution of the wine's quality.

# Bivariate Plots Section

```{r echo=FALSE, fig.align='center', Bivariate_Plots}
# How to do it for scatter plots?
ggplot(data = melt(data = subset(wine_df, select = -X), id.vars = 'quality'), mapping = aes(x = value, y = quality)) + 
  geom_jitter(color ='black', fill=I('#328484'), alpha=aes(0.002)) +
  facet_wrap(~variable, scales = 'free_x') +
  ylab('Quality') +
  xlab(label = '')
```

From the scatter plots, alcohol seems to be the most attribute that is related to the quality of the wine.<br>
Alcohol is the property that affects the brain cells, disrubts the consciousness of drinkers and make them feel happier.<br>
Thus, critics may prefer types of wine with higher alcohol contents.

```{r echo=FALSE, fig.align='center'}
ggplot(data = wine_df, mapping = aes(x=alcohol, y=quality)) +
  geom_jitter(alpha = aes(0.01), color=I('#328484'), size=2.5) +
  geom_smooth(method = 'lm', color=I('#889984')) +
  xlab('Alcohol quantity') +
  ylab('Quality')
```

Now, let's compute the correlation coefficients betweem the wine's quality and all the other attributes.

```{r}
cor(subset(wine_df, select = -c(X, quality)), y = wine_df$quality)
```

It is clear that there is a moderate correlation between the quality and the quantity of alcohol.<br>
Other correlation coefficients show weak correltaion coffecients or no correlation at all.


An obvious bivariate relation within the dataset is the free-total sulfur dioxide relation.
```{r echo=FALSE, fig.align='center'}
ggplot(data = wine_df, mapping = aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) +
  geom_point(alpha =aes(0.08)) +
  coord_cartesian(xlim = c(0, 100)) +
  geom_smooth(method = 'lm', color='black') +
  xlab('Free sulfur dioxide') +
  ylab('Total sulfur dioxide') + 
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))

```

```{r echo=FALSE, results='hide'}
cor(subset(wine_df, select = -c(X, quality)), y = wine_df$alcohol)
```

```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE, fig.align='center'}
ggplot(data = wine_df, mapping = aes(x = density, y = quality)) +
  geom_jitter(alpha = aes(0.05)) +
  geom_smooth(color= 'black') +
  coord_cartesian(xlim = c(0.99, 1.005))
```

# Bivariate Analysis
### Observations:
* There is a moderate correlation between the quality and the quantity of alcohol.
* There is an obvious bivariate relation within the dataset is the free-total sulfur dioxide relation.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(data = wine_df, mapping = aes(x=alcohol, y=quality)) +
  geom_jitter(aes(color=residual.sugar)) +
  scale_color_gradient2(low="blue", high="black") +
  geom_smooth(method = 'lm', color=I('#889984')) +
  xlab('Alcohol quantity') +
  ylab('Quality') +
  theme_dark()

fit <- lm(quality ~ residual.sugar + alcohol + density + pH + citric.acid + sulphates,
          data=wine_df)
summary(fit) # show results

wine_df$pred <- predict(fit, wine_df)

ggplot(data = wine_df, mapping = aes(x = pred, y = quality)) +
  geom_jitter(alpha = aes(0.02))
```

# Multivariate Analysis

### Observations:
* Residual sugar has no clear effect on the relation between quality and alocohol contents.
* The multiple regression model isn't effective for prediction of quality.

# Final Plots and Summary

### Plot One
```{r echo=FALSE, fig.align='center', Plot_One}
ggplot(data = wine_df, mapping = aes(x = factor(quality))) + 
  geom_bar(color ='black', fill=I('#328484')) +
  ylab('Count of wine samples') +
  xlab('Quality') +
  ggtitle('Distribution of wine\'s quality')
```

### Description One
The quality of white wine has a normal distribution with a mean value = 5.878 and standard deviation = 0.8856386.

### Plot Two
```{r echo=FALSE, results='hide', fig.align='center', warning=FALSE, message=FALSE, Plot_Two}
ggplot(data = wine_df, mapping = aes(x = density, y = quality)) +
  geom_jitter(alpha = aes(0.05)) +
  geom_smooth(color= 'black') +
  coord_cartesian(xlim = c(0.99, 1.005)) +
  xlab('Density (g / cm^3)') +
  ylab('Quality') +
  ggtitle('Effect of density on quality')
```

### Description Two
The plot shows the effect of wine's density on its quality.<br>
Wine with lower densities seem to get higher ratings.<br>
Increasing the density of wine might make it unpleasent to drinkers.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = wine_df, mapping = aes(x=alcohol, y=quality)) +
  geom_jitter(alpha = aes(0.01), color=I('#328484'), size=2.5) +
  geom_smooth(method = 'lm', color=I('#889984')) +
  xlab('Alcohol quantity (% by volume)') +
  ylab('Quality') +
  ggtitle('Effect of alcohol on quality')
```

### Description Three
The plot describes the effect of Alcohol on the wine's rating.<br>
As mentioned before, alcohol seems to be the most attribute that is related to the quality of the wine.<br>
Alcohol is the property that affects the brain cells, disrubts the consciousness of drinkers and make them feel happier.<br>
Thus, critics may prefer types of wine with higher alcohol contents.

# Reflection

The white wine dataset is challenging as most of the attributes are chemical.<br>
This fact makes intuitive explanations more difficult and requires deep understanding of these 
attributes on the wine's quality.<br>
More efforts could be done in exploring the inter-dependency between the different attributes.<br>
Plotting the histograms of all the attributes was a good way to understand their distributions.<br>
Additionally, Checking the correlation coefficients between quality and the rest of the attributes<br>
supported the rest of the investigation results.